---
title: "A primer on spatial quantile regression with autocorrelation"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: spmixqr.bib
vignette: >
  %\VignetteIndexEntry{A primer on spatial quantile regression with autocorrelation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4.2, dpi = 96)
has_gg <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r setup}
library(spmixqr)
set.seed(1)
```

This primer is for a researcher with *areal* spatial data (regions, neighbourhoods,
census tracts) who wants to know how covariates move the **tails** of an outcome,
not just its mean, while neighbouring units are spatially dependent. We work the
Columbus, Ohio neighbourhood-crime dataset from start to finish: building the spatial
weights, fitting a spatial quantile regression with a conditional-autoregressive (CAR)
error, reading the coefficients across the distribution, checking whether the spatial
term absorbs the autocorrelation, and reporting the result.

## 1. The question

The 49 Columbus neighbourhoods differ sharply in residential crime. Two questions an
ordinary (mean) regression cannot answer:

* Does household **income** protect against crime *equally* across the distribution, or
  does it matter more in the **high-crime** neighbourhoods (the upper tail)?
* After income and housing value are accounted for, do neighbouring areas still have
  **correlated** crime, a spatial *spillover* the model is missing, and can we absorb it?

The spatial-error (CAR) quantile regression in `spmixqr` answers both at once. It fits a
quantile regression at any level `tau` and adds a spatial random effect, built from a
neighbourhood **weights matrix**, that soaks up residual spatial dependence.

## 2. The model in one page

For a chosen quantile level $\tau$, the $\tau$-th conditional quantile of the outcome in
region $s$ is
$$ Q_\tau(y_s \mid x_s) = x_s'\beta + \phi_s, \qquad \phi \sim \text{CAR}(W,\alpha,\lambda_\phi). $$
Here $\beta$ are the covariate effects at quantile $\tau$, and $\phi$ is a spatial random
effect, one value per region, with a conditional-autoregressive precision built from the
spatial weights matrix $W$. We use the **Leroux** form
$Q(\alpha)=\alpha(D-W)+(1-\alpha)I$, where $D$ is the degree matrix: $\alpha$ controls
spatial strength and $\lambda_\phi$ its smoothness. We constrain $\phi$ to sum to zero
within each connected component, so the intercept carries the overall level and $\phi$ is
the mean-zero spatial *deviation*. The quantile fit uses the asymmetric-Laplace
(check-loss) working likelihood; estimation is a penalised EM in which the CAR precision
enters as a sparse ridge penalty [@kozumi2011; @leroux2000]. With one regime (`G = 1`,
used throughout this primer) the model is a single-population spatial-error quantile
regression; the mixture extension is covered in `vignette("spmixqr-mixtures")`.

## 3. The data

```{r data}
data(columbus_crime)
str(columbus_crime)
summary(columbus_crime$crime)
```

`crime` is residential burglaries and vehicle thefts per 1000 households; `income` and
`hoval` (housing value) are in \$1000; `x` and `y` are neighbourhood centroids. Crime
ranges from near zero to nearly 70 per 1000, a wide right-leaning spread that makes the
tails, not just the average, worth modelling [@anselin1988].

## 4. Building the spatial weights

`spq_weights()` constructs the weights matrix `W`. For areal data the natural choice is
**queen** (or **rook**) contiguity from the region polygons. We ship the queen contiguity
for Columbus as `columbus_W`, so we pass it through the `"supplied"` path.

```{r weights}
data(columbus_W)
W <- spq_weights(columbus_W, type = "supplied")
W
```

From your own `sf`/`sp` polygons you would instead write
`spq_weights(polys, type = "queen")` (or `"rook"`); for point data, use
`spq_weights(coords, type = "distance", d2 = ...)` or `type = "knn"`. All paths return a
symmetric `W`, and `spmixqr` forms the CAR precision internally.

## 5. Fitting your first model

We model the **median** ($\tau = 0.5$) of crime on income and housing value, with a CAR
spatial-error term on the queen graph. For areal CAR fits, set `spatial_coef = FALSE`
(the mgcv slope-surface machinery is for point covariate effects, not the areal CAR term)
and pass the region identifier as `coords`.

```{r fit}
reg <- factor(columbus_crime$id)
fit <- spmixqr(crime ~ income + hoval, data = columbus_crime, coords = reg,
               G = 1, tau = 0.5, spatial_error = TRUE, spatial_coef = FALSE,
               spatial_W = W, control = spmixqr_control(nstart = 1L, seed = 1))
fit
```

## 6. Interpreting the estimates

```{r summary}
summary(fit)
```

At the median, each extra \$1000 of neighbourhood income is associated with about two
fewer crimes per 1000 households (`income` $\approx -2.0$), and the association is firmly
significant. Housing value adds little once income is in the model: its coefficient is
small and not significant. The standard errors shown are the fast
**classification-conditional** sandwich, here the smoothed-quantile sandwich of the
single-population fit. But the asymmetric-Laplace is a *working* likelihood and the data
are spatially dependent, so the **spatial-block bootstrap** (Section 10) is what to
report.

## 7. The headline question: does the income effect vary across the distribution?

A mean regression gives one income slope. Quantile regression gives a slope at each
`tau`, which lets us see whether income matters more where crime is high. We refit across
the distribution and read off the income coefficient.

```{r path}
taus <- seq(0.15, 0.85, by = 0.1)
inc <- sapply(taus, function(t) {
  f <- spmixqr(crime ~ income + hoval, data = columbus_crime, coords = reg,
               G = 1, tau = t, spatial_error = TRUE, spatial_coef = FALSE,
               spatial_W = W, variance = "none",
               control = spmixqr_control(nstart = 1L, seed = 1))
  coef(f)["income", 1]
})
data.frame(tau = taus, income_coef = round(inc, 2))
```

The protective association **strengthens** from the lower to the upper part of the crime
distribution: income's coefficient is near $-1.5$ around the first quartile and steepens
to roughly $-2.2$ in the high-crime upper tail. Income differences matter *more* for
separating moderate- from high-crime neighbourhoods than for the quietest ones, a pattern
the conditional mean averages away.

## 8. Accounting for spatial autocorrelation

Does the model leave behind spatial structure, neighbouring areas with correlated
residual crime? `moran_resid()` computes a permutation Moran's I on the residuals
**before** and **after** the CAR term.

```{r moran}
mr <- moran_resid(fit, nsim = 999)
mr
```

Before the spatial term, the residuals show borderline positive clustering
(Moran's I $\approx 0.17$, $p \approx 0.05$): nearby neighbourhoods have similar
left-over crime. After the CAR term, the clustering drops to non-significant
($I \approx 0.15$, $p \approx 0.10$). The shift is modest but real: the spatial effect
absorbs part of the spillover, and what remains no longer clears the usual threshold. The
fitted $\phi$ are the estimated spatial deviations, the "residual hotspot/coldspot"
surface that income and housing value do not explain:

```{r phi}
ph <- phi_surface(fit)
head(ph)
range(round(ph$phi, 2))
```

## 9. Reading the fit in pictures

A map of the spatial effect shows where crime runs higher or lower than the covariates
predict. We plot neighbourhood centroids shaded by $\phi$; a choropleth needs the
polygons, but centroids tell the same story here.

```{r maps, eval = has_gg, fig.alt = "Map of the CAR spatial effect across Columbus neighbourhoods"}
library(ggplot2)
theme_set(theme_minimal(base_size = 11))
ph$x <- columbus_crime$x; ph$y <- columbus_crime$y
ggplot(ph, aes(x, y, colour = phi)) +
  geom_point(size = 4) +
  scale_colour_gradient2(low = "#2c7bb6", mid = "grey90", high = "#d7191c") +
  labs(title = "CAR spatial effect (residual crime not explained by income/hoval)",
       x = "easting", y = "northing", colour = expression(phi))
```

```{r path-plot, eval = has_gg, fig.alt = "Income coefficient across quantiles"}
ggplot(data.frame(tau = taus, income = inc), aes(tau, income)) +
  geom_line() + geom_point() + geom_hline(yintercept = 0, linetype = 3) +
  labs(title = "Income effect on crime across the distribution",
       x = expression(tau), y = "income coefficient")
```

The income-by-`tau` curve is the headline finding in one picture: the slope grows more
negative as we move up the crime distribution.

## 10. Diagnostics: can you trust it?

```{r diag}
d <- fit$diagnostics
c(converged = d$converged,
  moran_before_p = round(mr$before$p_value, 3),
  moran_after_p  = round(mr$after$p_value, 3))
```

* **Did the spatial term help?** Compare residual Moran's I before and after (Section 8).
  If it was already non-significant, there is little spatial structure to absorb: the
  "negative control" case.
* **Inference.** Report the **spatial-block bootstrap**, not the sandwich:
  `spmixqr(..., variance = "boot")` resamples connected blocks of the graph and refits.
* **Weak identifiability.** The CAR strength `car_alpha` is held fixed (default 0.95), and
  $\lambda_\phi$ is selectable by BIC via `spmixqr_select()`. Neither is a precisely
  estimable variance, so we do not over-interpret their magnitudes.
* **Small samples.** Columbus has 49 units. The point estimates are stable, but spatial
  variance parameters are weakly identified at this size; lean on the bootstrap and avoid
  many regimes.

## 11. Practical guidance and pitfalls

* **Areal CAR fits use `spatial_coef = FALSE`** and pass the region id as `coords`. The
  CAR term carries the spatial level; the slope-surface machinery (`spatial_coef = TRUE`)
  is the separate point-data feature in `vignette("spmixqr-mixtures")`.
* **Choosing `W`.** Queen contiguity is the default for polygons; rook is stricter
  (shared edges only); for point data, use distance bands or k-NN. Check the substantive
  conclusions against a second `W`.
* **Gate vs CAR.** A free spatial *gate* and a CAR *intercept* are not separately
  identified. Setting `spatial_error = TRUE` turns the spatial gate off, and an explicit
  `spatial_gate = TRUE` raises a guardrail error. A covariate gate (`gating = ~ z`) is
  fine.
* **Regimes.** `G > 1` gives a mixture of spatial quantile regressions (latent
  neighbourhood types); use it only with enough units.

## 12. How `spmixqr` relates to other tools

`quantreg` fits quantile regression with no spatial structure; `spatialreg` and `spdep`
fit spatial autoregressive *mean* models; `CARBayes` fits CAR models but not a quantile
likelihood. The only dedicated spatial quantile package, `McSpatial`, has been archived
since 2021. `spmixqr` provides a maintained spatial-error quantile regression with
first-class queen, rook, distance, k-NN, and user-supplied weights, a Moran's-I
diagnostic, and the mixture extension [@reich2011].

## 13. Reporting and reproducibility

```{r report}
list(tau = fit$tau,
     income = round(coef(fit)["income", 1], 2),
     hoval  = round(coef(fit)["hoval", 1], 2),
     moran_before_p = round(mr$before$p_value, 3),
     moran_after_p  = round(mr$after$p_value, 3))
```

A spatial-error quantile regression turns "how does income relate to crime" into a
sharper, map-aware answer: income's protective association is strongest in the high-crime
tail, and once a CAR term absorbs the residual neighbourhood spillover, the leftover
spatial clustering no longer clears the usual significance threshold.

## References
