---
title: "Getting started with spmixqr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with spmixqr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

`spmixqr` fits **spatial finite mixtures of quantile regressions**. The data are
partitioned into a few latent *regimes*. Each regime is a quantile regression, the
probability of belonging to a regime varies smoothly over space, and each regime's
covariate slopes are themselves spatial surfaces (the per-regime intercept stays
scalar). With one regime it is a penalised spatially-varying-coefficient quantile
regression.

```{r setup}
library(spmixqr)
set.seed(1)
```

## A first fit

We simulate two regimes whose mixing probability drifts across space and whose slope
on `x` varies with location, then recover them.

```{r fit}
d <- sim_spmixqr(n = 250, G = 2, tau = 0.5, seed = 1)
fit <- spmixqr(y ~ x, data = d$data, coords = d$coords, G = 2, tau = 0.5,
               control = spmixqr_control(nstart = 4L, seed = 1))
fit
```

## Reading the spatial structure

The gate maps where each regime is likely; `plot()` shows it over the coordinates.

```{r gate, fig.alt = "Gate probability map"}
plot(fit, which = "gate", regime = 2)
```

The recovered membership tracks the truth:

```{r recover}
cl <- predict(fit, type = "class")
acc <- max(mean(cl == d$truth$z), mean((3 - cl) == d$truth$z))
round(acc, 2)
```

See `vignette("spmixqr-primer")` for a full applied walk-through on real data,
including interpretation, uncertainty, and diagnostics.
