---
title: "A primer on spatial quantile regression with autocorrelation"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: spmixqr.bib
vignette: >
  %\VignetteIndexEntry{A primer on spatial quantile regression with autocorrelation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4.2, dpi = 96)
has_gg  <- requireNamespace("ggplot2", quietly = TRUE)
has_sf  <- requireNamespace("sf", quietly = TRUE)
has_map <- has_gg && has_sf
```

```{r setup}
library(spmixqr)
set.seed(1)
```

This primer is for a researcher with *areal* spatial data (regions, neighbourhoods,
census tracts) who wants to know how covariates move the **tails** of an outcome,
not just its mean, while neighbouring units are spatially dependent. We work the
Columbus, Ohio neighbourhood-crime dataset from start to finish, in two examples:

* **Example A** — a single-population spatial-error (CAR) quantile regression: build the
  weights, fit, read the coefficients across the distribution, and absorb the residual
  spatial autocorrelation.
* **Example B** — a two-regime **mixture** of quantile regressions: uncover latent
  neighbourhood crime regimes and map them.

Throughout we start from a **shapefile**, read it as an `sf` map, and put the estimates
and residuals back on the map as choropleths alongside the usual tables.

## 1. The question

The 49 Columbus neighbourhoods differ sharply in residential crime. Two questions an
ordinary (mean) regression cannot answer:

* Does household **income** protect against crime *equally* across the distribution, or
  does it matter more in the **high-crime** neighbourhoods (the upper tail)?
* After income and housing value are accounted for, do neighbouring areas still have
  **correlated** crime, a spatial *spillover* the model is missing, and can we absorb it?

The spatial-error (CAR) quantile regression in `spmixqr` answers both at once. It fits a
quantile regression at any level `tau` and adds a spatial random effect, built from a
neighbourhood **weights matrix**, that soaks up residual spatial dependence.

## 2. The model in one page

For a chosen quantile level $\tau$, the $\tau$-th conditional quantile of the outcome in
region $s$ is
$$ Q_\tau(y_s \mid x_s) = x_s'\beta + \phi_s, \qquad \phi \sim \text{CAR}(W,\alpha,\lambda_\phi). $$
Here $\beta$ are the covariate effects at quantile $\tau$, and $\phi$ is a spatial random
effect, one value per region, with a conditional-autoregressive precision built from the
spatial weights matrix $W$. We use the **Leroux** form
$Q(\alpha)=\alpha(D-W)+(1-\alpha)I$, where $D$ is the degree matrix: $\alpha$ controls
spatial strength and $\lambda_\phi$ its smoothness. We constrain $\phi$ to sum to zero
within each connected component, so the intercept carries the overall level and $\phi$ is
the mean-zero spatial *deviation*. The quantile fit uses the asymmetric-Laplace
(check-loss) working likelihood; estimation is a penalised EM in which the CAR precision
enters as a sparse ridge penalty [@kozumi2011; @leroux2000]. With one regime (`G = 1`,
Example A) the model is a single-population spatial-error quantile regression; with
`G > 1` (Example B) it becomes a mixture of quantile regressions with latent regimes.
The full mixture and spatially-varying-slope feature set is in
`vignette("spmixqr-mixtures")`.

## 3. The data

```{r data}
data(columbus_crime)
str(columbus_crime)
summary(columbus_crime$crime)
```

`crime` is residential burglaries and vehicle thefts per 1000 households; `income` and
`hoval` (housing value) are in \$1000; `x` and `y` are neighbourhood centroids. Crime
ranges from near zero to nearly 70 per 1000, a wide right-leaning spread that makes the
tails, not just the average, worth modelling [@anselin1988].

## 4. Building the spatial weights

`spq_weights()` constructs the weights matrix `W`. For areal data the natural choice is
**queen** (or **rook**) contiguity from the region polygons. We ship the queen contiguity
for Columbus as `columbus_W`, so we pass it through the `"supplied"` path.

```{r weights}
data(columbus_W)
W <- spq_weights(columbus_W, type = "supplied")
W
```

From your own `sf`/`sp` polygons you would instead write
`spq_weights(polys, type = "queen")` (or `"rook"`); for point data, use
`spq_weights(coords, type = "distance", d2 = ...)` or `type = "knn"`. All paths return a
symmetric `W`, and `spmixqr` forms the CAR precision internally.

## 5. Fitting your first model

We model the **median** ($\tau = 0.5$) of crime on income and housing value, with a CAR
spatial-error term on the queen graph. For areal CAR fits, set `spatial_coef = FALSE`
(the mgcv slope-surface machinery is for point covariate effects, not the areal CAR term)
and pass the region identifier as `coords`.

```{r fit}
reg <- factor(columbus_crime$id)
fit <- spmixqr(crime ~ income + hoval, data = columbus_crime, coords = reg,
               G = 1, tau = 0.5, spatial_error = TRUE, spatial_coef = FALSE,
               spatial_W = W, control = spmixqr_control(nstart = 1L, seed = 1))
fit
```

## 6. Interpreting the estimates

```{r summary}
summary(fit)
```

At the median, each extra \$1000 of neighbourhood income is associated with about two
fewer crimes per 1000 households (`income` $\approx -2.0$), and the association is firmly
significant. Housing value adds little once income is in the model: its coefficient is
small and not significant. The standard errors shown are the fast
**classification-conditional** sandwich, here the smoothed-quantile sandwich of the
single-population fit. But the asymmetric-Laplace is a *working* likelihood and the data
are spatially dependent, so the **spatial-block bootstrap** (Section 10) is what to
report.

## 7. The headline question: does the income effect vary across the distribution?

A mean regression gives one income slope. Quantile regression gives a slope at each
`tau`, which lets us see whether income matters more where crime is high. We refit across
the distribution and read off the income coefficient.

```{r path}
taus <- seq(0.15, 0.85, by = 0.1)
inc <- sapply(taus, function(t) {
  f <- spmixqr(crime ~ income + hoval, data = columbus_crime, coords = reg,
               G = 1, tau = t, spatial_error = TRUE, spatial_coef = FALSE,
               spatial_W = W, variance = "none",
               control = spmixqr_control(nstart = 1L, seed = 1))
  coef(f)["income", 1]
})
data.frame(tau = taus, income_coef = round(inc, 2))
```

The protective association **strengthens** from the lower to the upper part of the crime
distribution: income's coefficient is near $-1.5$ around the first quartile and steepens
to roughly $-2.2$ in the high-crime upper tail. Income differences matter *more* for
separating moderate- from high-crime neighbourhoods than for the quietest ones, a pattern
the conditional mean averages away.

## 8. Accounting for spatial autocorrelation

Does the model leave behind spatial structure, neighbouring areas with correlated
residual crime? `moran_resid()` computes a permutation Moran's I on the residuals
**before** and **after** the CAR term.

```{r moran}
mr <- moran_resid(fit, nsim = 999)
mr
```

Before the spatial term, the residuals show borderline positive clustering
(Moran's I $\approx 0.17$, $p \approx 0.05$): nearby neighbourhoods have similar
left-over crime. After the CAR term, the clustering drops to non-significant
($I \approx 0.15$, $p \approx 0.10$). The shift is modest but real: the spatial effect
absorbs part of the spillover, and what remains no longer clears the usual threshold. The
fitted $\phi$ are the estimated spatial deviations, the "residual hotspot/coldspot"
surface that income and housing value do not explain:

```{r phi}
ph <- phi_surface(fit)
head(ph)
range(round(ph$phi, 2))
```

## 9. Reading the fit on the map

Estimates and residuals mean more on a map than in a column. We start from the
neighbourhood **shapefile** (shipped with the package as a GeoPackage), read it as an
`sf` object, and attach the fit's quantities. The polygons are in the same order as
`columbus_crime`, so a column bind aligns them.

```{r sf-read, eval = has_sf, message = FALSE}
library(sf)
shp <- system.file("shapes/columbus.gpkg", package = "spmixqr")
cols <- st_read(shp, quiet = TRUE)               # 49 neighbourhood polygons
st_crs(cols) <- NA                               # Columbus coords are arbitrary planar units
cols$crime <- columbus_crime$crime
cols$phi   <- phi_surface(fit)$phi               # the CAR spatial effect, per region
cols$resid <- residuals(fit)                     # observed minus fitted median crime
```

A choropleth of $\phi$ shows where crime runs higher (red) or lower (blue) than income
and housing value predict, after the spatial smoothing. A second map of the residuals
shows what the whole model still misses.

```{r map-phi, eval = has_map, fig.alt = "Choropleth of the CAR spatial effect across Columbus"}
library(ggplot2)
theme_set(theme_void(base_size = 11))
ggplot(cols) +
  geom_sf(aes(fill = phi), colour = "grey40", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2c7bb6", mid = "grey92", high = "#d7191c") +
  labs(title = "CAR spatial effect on crime",
       subtitle = "residual neighbourhood level not explained by income or housing value",
       fill = expression(phi))
```

```{r map-resid, eval = has_map, fig.alt = "Choropleth of model residuals across Columbus"}
ggplot(cols) +
  geom_sf(aes(fill = resid), colour = "grey40", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2c7bb6", mid = "grey92", high = "#d7191c") +
  labs(title = "Median-model residuals", fill = "resid")
```

The $\phi$ map localises the spillover the Moran's I test flagged: a cluster of
neighbourhoods whose crime sits above what their income and housing value imply, which
the CAR term pools toward its neighbours. The income-by-`tau` curve puts the headline
finding in one picture.

```{r path-plot, eval = has_gg, fig.width = 5, fig.height = 3.5, fig.alt = "Income coefficient across quantiles"}
ggplot(data.frame(tau = taus, income = inc), aes(tau, income)) +
  geom_line() + geom_point() + geom_hline(yintercept = 0, linetype = 3) +
  theme_minimal(base_size = 11) +
  labs(title = "Income effect on crime across the distribution",
       x = expression(tau), y = "income coefficient")
```

The slope grows more negative as we move up the crime distribution.

## 10. Example B: a mixture of crime regimes

So far one relationship held for every neighbourhood. A **mixture** (`G = 2`) instead
lets the data split into latent *regimes* — neighbourhood types with their own
crime/income relationship — and assigns each area a posterior probability of belonging
to each. Here we fit two regimes with a constant gate (no spatial term), which keeps
the example self-contained; a covariate or spatial gate is the next step.

```{r mixfit}
fit2 <- spmixqr(crime ~ income + hoval, data = columbus_crime, G = 2, tau = 0.5,
                spatial_gate = FALSE, spatial_coef = FALSE,
                control = spmixqr_control(nstart = 8L, seed = 1))
round(coef(fit2), 2)        # one column per regime
table(regime = predict(fit2, type = "class"))
```

The two regimes are distinct. One (the smaller group) has a **high baseline** crime
level and a **steep** income gradient — income near $-2.8$ — so income separates
neighbourhoods sharply there. The other (the larger group) has a **lower baseline** and
a **gentler** income gradient (near $-1.4$). The mixture has found that the
income–crime relationship is not one curve but two.

Mapping the regime each neighbourhood is assigned to shows the split is spatially
coherent even though the gate used no location information: the high-gradient regime
concentrates rather than scattering at random.

```{r map-regime, eval = has_map, fig.alt = "Choropleth of mixture regime membership across Columbus"}
cols$regime <- factor(predict(fit2, type = "class"))
cols$p2     <- predict(fit2, type = "posterior")[, 2]
ggplot(cols) +
  geom_sf(aes(fill = regime), colour = "grey40", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Assigned crime regime", fill = "regime")
```

```{r map-resid2, eval = has_map, fig.alt = "Choropleth of mixture-model residuals across Columbus"}
cols$resid2 <- residuals(fit2)
ggplot(cols) +
  geom_sf(aes(fill = resid2), colour = "grey40", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2c7bb6", mid = "grey92", high = "#d7191c") +
  labs(title = "Mixture-model residuals", fill = "resid")
```

The residual map is the reality check: a mixture can fit closely in-sample, so look for
spatial pattern left in the residuals. With only 49 units two regimes is about as far as
the data support; for a spatially *driven* gate, or regime-specific spatial-error terms,
you need more units (and `vignette("spmixqr-mixtures")` covers the gate machinery).

## 11. Diagnostics: can you trust it?

```{r diag}
d <- fit$diagnostics
c(converged = d$converged,
  moran_before_p = round(mr$before$p_value, 3),
  moran_after_p  = round(mr$after$p_value, 3))
```

* **Did the spatial term help?** Compare residual Moran's I before and after (Section 8).
  If it was already non-significant, there is little spatial structure to absorb: the
  "negative control" case.
* **Inference.** Report the **spatial-block bootstrap**, not the sandwich:
  `spmixqr(..., variance = "boot")` resamples connected blocks of the graph and refits.
* **Weak identifiability.** The CAR strength `car_alpha` is held fixed (default 0.95), and
  $\lambda_\phi$ is selectable by BIC via `spmixqr_select()`. Neither is a precisely
  estimable variance, so we do not over-interpret their magnitudes.
* **Small samples.** Columbus has 49 units. The point estimates are stable, but spatial
  variance parameters are weakly identified at this size; lean on the bootstrap and avoid
  many regimes.

## 12. Practical guidance and pitfalls

* **Areal CAR fits use `spatial_coef = FALSE`** and pass the region id as `coords`. The
  CAR term carries the spatial level; the slope-surface machinery (`spatial_coef = TRUE`)
  is the separate point-data feature in `vignette("spmixqr-mixtures")`.
* **Choosing `W`.** Queen contiguity is the default for polygons; rook is stricter
  (shared edges only); for point data, use distance bands or k-NN. Check the substantive
  conclusions against a second `W`.
* **Gate vs CAR.** A free spatial *gate* and a CAR *intercept* are not separately
  identified. Setting `spatial_error = TRUE` turns the spatial gate off, and an explicit
  `spatial_gate = TRUE` raises a guardrail error. A covariate gate (`gating = ~ z`) is
  fine.
* **Regimes.** `G > 1` gives a mixture of spatial quantile regressions (latent
  neighbourhood types); use it only with enough units.

## 13. How `spmixqr` relates to other tools

`quantreg` fits quantile regression with no spatial structure; `spatialreg` and `spdep`
fit spatial autoregressive *mean* models; `CARBayes` fits CAR models but not a quantile
likelihood. The only dedicated spatial quantile package, `McSpatial`, has been archived
since 2021. `spmixqr` provides a maintained spatial-error quantile regression with
first-class queen, rook, distance, k-NN, and user-supplied weights, a Moran's-I
diagnostic, and the mixture extension [@reich2011].

## 14. Reporting and reproducibility

```{r report}
list(tau = fit$tau,
     income = round(coef(fit)["income", 1], 2),
     hoval  = round(coef(fit)["hoval", 1], 2),
     moran_before_p = round(mr$before$p_value, 3),
     moran_after_p  = round(mr$after$p_value, 3))
```

A spatial-error quantile regression turns "how does income relate to crime" into a
sharper, map-aware answer: income's protective association is strongest in the high-crime
tail, and once a CAR term absorbs the residual neighbourhood spillover, the leftover
spatial clustering no longer clears the usual significance threshold.

## References
