---
title: "Getting started with spmixqr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with spmixqr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

`spmixqr` fits **quantile regression for spatial data while accounting for spatial
autocorrelation**. A conditional-autoregressive (CAR) spatial random effect, built
from a weights matrix, absorbs residual spatial dependence; a Moran's I diagnostic
checks whether any spatial structure is left over. It extends to finite mixtures of
spatial quantile regressions.

```{r setup}
library(spmixqr)
```

## A first fit

We model the median of Columbus neighbourhood crime on income and housing value, with
a CAR spatial-error term on the queen-contiguity graph.

```{r fit}
data(columbus_crime); data(columbus_W)
W <- spq_weights(columbus_W, type = "supplied")

fit <- spmixqr(crime ~ income + hoval, data = columbus_crime,
               coords = factor(columbus_crime$id), G = 1, tau = 0.5,
               spatial_error = TRUE, spatial_coef = FALSE, spatial_W = W,
               control = spmixqr_control(nstart = 1L, seed = 1))
coef(fit)
```

Higher neighbourhood income is associated with substantially less crime at the median.

## Did the spatial term absorb the autocorrelation?

```{r moran}
moran_resid(fit)
```

Residual spatial clustering that is present before the CAR term drops toward
non-significance after it: the spatial effect soaks up the neighbourhood spillover.

## Observation weights

Pass a per-observation `weights` vector (a survey/sampling weight, a frequency count,
or a precision weight) to weight the fit. This is distinct from the spatial neighbour
matrix `spatial_W`. `weights_type` chooses the interpretation: it does **not** change
the point estimates, only the standard errors and the effective sample size.

```{r weights}
set.seed(1)
w <- runif(nrow(columbus_crime), 0.5, 2)          # e.g. inverse inclusion probabilities

fit_w <- spmixqr(crime ~ income + hoval, data = columbus_crime,
                 coords = factor(columbus_crime$id), G = 1, tau = 0.5,
                 spatial_error = TRUE, spatial_coef = FALSE, spatial_W = W,
                 weights = w, weights_type = "sampling",
                 control = spmixqr_control(nstart = 1L, seed = 1))
coef(fit_w)
```

Use `"sampling"` for inverse-inclusion / survey weights (design-robust SEs),
`"frequency"` for integer replication counts (SEs and AIC/BIC then reflect the total
count `sum(weights)`), or `"precision"` for analytic `1/variance` weights. The
spatial-block bootstrap (`variance = "boot"`) is the recommended inference path and
is weight-aware for every type.

See `vignette("spmixqr-primer")` for the full worked example (how income's effect
varies across the crime distribution, mapping the spatial effect, inference, and
diagnostics) and `vignette("spmixqr-mixtures")` for the finite-mixture features.
