---
title: "A primer on spatial quantile regression with autocorrelation"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: spmixqr.bib
vignette: >
  %\VignetteIndexEntry{A primer on spatial quantile regression with autocorrelation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4.2, dpi = 96)
has_gg  <- requireNamespace("ggplot2", quietly = TRUE)
has_sf  <- requireNamespace("sf", quietly = TRUE)
has_map <- has_gg && has_sf
```

```{r setup}
library(spmixqr)
set.seed(1)
```

This primer is for a researcher with *areal* spatial data (regions, neighbourhoods,
census tracts) who wants to know how covariates move the **tails** of an outcome,
not just its mean, while neighbouring units are spatially dependent. We work the
Columbus, Ohio neighbourhood-crime dataset from start to finish, in two examples:

* **Example A** — a single-population spatial-error (CAR) quantile regression: build the
  weights, fit, read the coefficients across the distribution, and absorb the residual
  spatial autocorrelation.
* **Example B** — a two-regime **mixture** of quantile regressions on a larger dataset
  (Lucas County, Ohio house sales): uncover latent housing *submarkets* with a spatial
  gate and map them.

In Example A we start from a **shapefile**, read it as an `sf` map, and put the estimates
and residuals back on the map as choropleths alongside the tables; Example B does the
same on point geometry. A mixture needs more units than the 49 Columbus neighbourhoods,
which is why it uses the larger housing data.

## 1. The question

The 49 Columbus neighbourhoods differ sharply in residential crime. Two questions an
ordinary (mean) regression cannot answer:

* Does household **income** protect against crime *equally* across the distribution, or
  does it matter more in the **high-crime** neighbourhoods (the upper tail)?
* After income and housing value are accounted for, do neighbouring areas still have
  **correlated** crime, a spatial *spillover* the model is missing, and can we absorb it?

The spatial-error (CAR) quantile regression in `spmixqr` answers both at once. It fits a
quantile regression at any level `tau` and adds a spatial random effect, built from a
neighbourhood **weights matrix**, that soaks up residual spatial dependence.

## 2. The model in one page

For a chosen quantile level $\tau$, the $\tau$-th conditional quantile of the outcome in
region $s$ is
$$ Q_\tau(y_s \mid x_s) = x_s'\beta + \phi_s, \qquad \phi \sim \text{CAR}(W,\alpha,\lambda_\phi). $$
Here $\beta$ are the covariate effects at quantile $\tau$, and $\phi$ is a spatial random
effect, one value per region, with a conditional-autoregressive precision built from the
spatial weights matrix $W$. We use the **Leroux** form
$Q(\alpha)=\alpha(D-W)+(1-\alpha)I$, where $D$ is the degree matrix: $\alpha$ controls
spatial strength and $\lambda_\phi$ its smoothness. We constrain $\phi$ to sum to zero
within each connected component, so the intercept carries the overall level and $\phi$ is
the mean-zero spatial *deviation*. The quantile fit uses the asymmetric-Laplace
(check-loss) working likelihood; estimation is a penalised EM in which the CAR precision
enters as a sparse ridge penalty [@kozumi2011; @leroux2000]. With one regime (`G = 1`,
Example A) the model is a single-population spatial-error quantile regression; with
`G > 1` (Example B) it becomes a mixture of quantile regressions with latent regimes.
The full mixture and spatially-varying-slope feature set is in
`vignette("spmixqr-mixtures")`.

## 3. The data

```{r data}
data(columbus_crime)
str(columbus_crime)
summary(columbus_crime$crime)
```

`crime` is residential burglaries and vehicle thefts per 1000 households; `income` and
`hoval` (housing value) are in \$1000; `x` and `y` are neighbourhood centroids. Crime
ranges from near zero to nearly 70 per 1000, a wide right-leaning spread that makes the
tails, not just the average, worth modelling [@anselin1988].

## 4. Building the spatial weights

`spq_weights()` constructs the weights matrix `W`. For areal data the natural choice is
**queen** (or **rook**) contiguity from the region polygons. We ship the queen contiguity
for Columbus as `columbus_W`, so we pass it through the `"supplied"` path.

```{r weights}
data(columbus_W)
W <- spq_weights(columbus_W, type = "supplied")
W
```

From your own `sf`/`sp` polygons you would instead write
`spq_weights(polys, type = "queen")` (or `"rook"`); for point data, use
`spq_weights(coords, type = "distance", d2 = ...)` or `type = "knn"`. All paths return a
symmetric `W`, and `spmixqr` forms the CAR precision internally.

## 5. Fitting your first model

We model the **median** ($\tau = 0.5$) of crime on income and housing value, with a CAR
spatial-error term on the queen graph. For areal CAR fits, set `spatial_coef = FALSE`
(the mgcv slope-surface machinery is for point covariate effects, not the areal CAR term)
and pass the region identifier as `coords`.

```{r fit}
reg <- factor(columbus_crime$id)
fit <- spmixqr(crime ~ income + hoval, data = columbus_crime, coords = reg,
               G = 1, tau = 0.5, spatial_error = TRUE, spatial_coef = FALSE,
               spatial_W = W, variance = "boot",
               control = spmixqr_control(nstart = 1L, seed = 1, boot_B = 60L))
fit
```

```{r boot-note, echo = FALSE, results = "asis"}
cat("> A small `boot_B` keeps this vignette fast; use a few hundred replicates for reporting.")
```

## 6. Interpreting the estimates

```{r summary}
summary(fit)
```

At the median, each extra \$1000 of neighbourhood income is associated with about two
fewer crimes per 1000 households (`income` $\approx -2.0$), and the association is firmly
significant. Housing value adds little once income is in the model: its coefficient is
small and not significant. These standard errors come from the **spatial-block
bootstrap** (`variance = "boot"`), the right choice for spatially dependent data under
the asymmetric-Laplace working likelihood; a fast classification-conditional sandwich
(`variance = "sandwich"`) is also available.

## 7. The headline question: does the income effect vary across the distribution?

A mean regression gives one income slope. Quantile regression gives a slope at each
`tau`, which lets us see whether income matters more where crime is high. We refit across
the distribution and read off the income coefficient.

```{r path}
taus <- seq(0.15, 0.85, by = 0.1)
inc <- sapply(taus, function(t) {
  f <- spmixqr(crime ~ income + hoval, data = columbus_crime, coords = reg,
               G = 1, tau = t, spatial_error = TRUE, spatial_coef = FALSE,
               spatial_W = W, variance = "none",
               control = spmixqr_control(nstart = 1L, seed = 1))
  coef(f)["income", 1]
})
data.frame(tau = taus, income_coef = round(inc, 2))
```

The protective association **strengthens** from the lower to the upper part of the crime
distribution: income's coefficient is near $-1.5$ around the first quartile and steepens
to roughly $-2.2$ in the high-crime upper tail. Income differences matter *more* for
separating moderate- from high-crime neighbourhoods than for the quietest ones, a pattern
the conditional mean averages away.

## 8. Accounting for spatial autocorrelation

Does the model leave behind spatial structure, neighbouring areas with correlated
residual crime? `moran_resid()` computes a permutation Moran's I on the residuals
**before** and **after** the CAR term.

```{r moran}
mr <- moran_resid(fit, nsim = 999)
mr
```

Before the spatial term, the residuals show borderline positive clustering
(Moran's I $\approx 0.17$, $p \approx 0.05$): nearby neighbourhoods have similar
left-over crime. After the CAR term, the clustering drops to non-significant
($I \approx 0.15$, $p \approx 0.10$). The shift is modest but real: the spatial effect
absorbs part of the spillover, and what remains no longer clears the usual threshold. The
fitted $\phi$ are the estimated spatial deviations, the "residual hotspot/coldspot"
surface that income and housing value do not explain. **The map alone is not enough**:
a roughness-penalised surface always looks smooth, so we need to know which deviations
are reliable. `phi_surface(ci = TRUE)` attaches bootstrap standard errors and flags the
`credible` units whose interval excludes zero.

```{r phi}
ph <- phi_surface(fit, ci = TRUE)              # adds se, lower, upper, credible
head(round(ph[, c("phi", "se", "lower", "upper")], 2))
sprintf("%d of %d neighbourhoods are credible hot/cold spots", sum(ph$credible), nrow(ph))
```

## 9. Reading the fit on the map

Estimates and residuals mean more on a map than in a column. We start from the
neighbourhood **shapefile** (shipped with the package as a GeoPackage), read it as an
`sf` object, and attach the fit's quantities. The polygons are in the same order as
`columbus_crime`, so a column bind aligns them.

```{r sf-read, eval = has_sf, message = FALSE}
library(sf)
shp <- system.file("shapes/columbus.gpkg", package = "spmixqr")
cols <- st_read(shp, quiet = TRUE)               # 49 neighbourhood polygons
st_crs(cols) <- NA                               # Columbus coords are arbitrary planar units
ps         <- phi_surface(fit, ci = TRUE)        # CAR effect + uncertainty, per region
cols$crime <- columbus_crime$crime
cols$phi   <- ps$phi                             # the CAR spatial effect
cols$credible_phi <- ifelse(ps$credible, ps$phi, NA)   # keep only reliable hot/cold spots
cols$resid <- residuals(fit)                     # observed minus fitted median crime
```

A choropleth of $\phi$ shows where crime runs higher (red) or lower (blue) than income
and housing value predict. But the interpretable map is the **credible** one: the same
surface with the unreliable cells greyed out, so the reader sees only the hot/cold spots
the data actually support. A residual map shows what the whole model still misses.

```{r map-phi, eval = has_map, fig.alt = "Choropleth of the CAR spatial effect across Columbus"}
library(ggplot2)
theme_set(theme_void(base_size = 11))
ggplot(cols) +
  geom_sf(aes(fill = phi), colour = "grey40", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2c7bb6", mid = "grey92", high = "#d7191c") +
  labs(title = "CAR spatial effect on crime",
       subtitle = "residual neighbourhood level (crimes per 1000) beyond income and housing",
       fill = expression(phi))
```

```{r map-credible, eval = has_map, fig.alt = "Credible CAR hot/cold spots across Columbus"}
ggplot(cols) +
  geom_sf(aes(fill = credible_phi), colour = "grey40", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2c7bb6", mid = "grey92", high = "#d7191c",
                       na.value = "grey92") +
  labs(title = "Credible hot/cold spots only",
       subtitle = "grey = interval includes zero (not reliably different from expected)",
       fill = expression(phi))
```

Read a credible $\phi$ in the outcome's own units: a value of, say, $+8$ means the
median crime rate there is about eight crimes per 1000 households **above** what income
and housing value predict, reliably so. (When the outcome is modelled on the log scale,
`phi_surface(scale = "exp")` instead reports `exp(phi)` as a multiplier --- 1.18 reads as
"about 18% above expected" --- which the `vignette("spmixqr-spatial-error")` shows.)

```{r map-resid, eval = has_map, fig.alt = "Choropleth of model residuals across Columbus"}
ggplot(cols) +
  geom_sf(aes(fill = resid), colour = "grey40", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2c7bb6", mid = "grey92", high = "#d7191c") +
  labs(title = "Median-model residuals", fill = "resid")
```

The $\phi$ map localises the spillover the Moran's I test flagged: a cluster of
neighbourhoods whose crime sits above what their income and housing value imply, which
the CAR term pools toward its neighbours. The income-by-`tau` curve puts the headline
finding in one picture.

```{r path-plot, eval = has_gg, fig.width = 5, fig.height = 3.5, fig.alt = "Income coefficient across quantiles"}
ggplot(data.frame(tau = taus, income = inc), aes(tau, income)) +
  geom_line() + geom_point() + geom_hline(yintercept = 0, linetype = 3) +
  theme_minimal(base_size = 11) +
  labs(title = "Income effect on crime across the distribution",
       x = expression(tau), y = "income coefficient")
```

The slope grows more negative as we move up the crime distribution.

## 10. Example B: a mixture of housing submarkets

A mixture needs enough data to resolve distinct regimes, so for this example we turn to
a larger dataset: roughly two thousand **house sales in Lucas County, Ohio** (a seeded
subsample of the classic real-estate data, shipped as `lucas_house`). The question is
whether a single price equation fits the whole county, or whether there are **submarkets**
— groups of homes whose price responds differently to size and age — and where they are.
These are *point* observations, so we use point coordinates rather than polygons.

```{r lucas-data}
data(lucas_house)
lucas <- transform(lucas_house, lprice = log(price), ltla = log(tla))
nrow(lucas)
```

A `G = 2` mixture lets the data split into two latent regimes. We give the mixing a
**spatial gate** (`spatial_gate = TRUE`), so the probability of belonging to each
submarket is a smooth surface over the county rather than a constant. BIC prefers the
mixture decisively over a single equation here.

```{r lucas-fit}
coords <- as.matrix(lucas[, c("x", "y")])
fitm <- spmixqr(lprice ~ ltla + age, data = lucas, coords = coords,
                G = 2, tau = 0.5, spatial_gate = TRUE, spatial_coef = FALSE,
                control = spmixqr_control(nstart = 4L, k = 20L, seed = 1))
round(coef(fitm), 2)          # one column per submarket
```

Both regimes share a price–size elasticity near 0.9 (a 1% larger home costs about 0.9%
more), but they differ in how steeply price falls with age: one submarket discounts age
**mildly** (age coefficient near $-0.8$), the other **steeply** (near $-1.7$, on the
per-century age scale of these data). The mixture has found two price processes where a
single regression would impose one.

The spatial gate makes the probability of the steep-discount submarket a surface. We map
that surface, the assigned submarket, and the residuals on the point geometry.

```{r lucas-sf, eval = has_sf, message = FALSE}
lucas_sf <- st_as_sf(lucas, coords = c("x", "y"))
lucas_sf$p_steep <- gate_surface(fitm)$prob[gate_surface(fitm)$regime == "2"]
lucas_sf$regime  <- factor(apply(predict(fitm, type = "posterior"), 1, which.max))
lucas_sf$resid   <- residuals(fitm)
```

```{r lucas-gate, eval = has_map, fig.alt = "Map of the spatial gate probability across Lucas County"}
ggplot(lucas_sf) +
  geom_sf(aes(colour = p_steep), size = 0.6) +
  scale_colour_viridis_c(limits = c(0, 1)) +
  labs(title = "Spatial gate: P(steep age-discount submarket)", colour = "prob")
```

```{r lucas-regime, eval = has_map, fig.alt = "Map of assigned housing submarket across Lucas County"}
ggplot(lucas_sf) +
  geom_sf(aes(colour = regime), size = 0.6) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "Assigned submarket", colour = "regime")
```

The gate surface localises the steep-discount submarket to part of the county rather than
scattering it at random — the mixing is genuinely spatial. The residual map is the
reality check: a flexible mixture can fit closely in-sample, so look for spatial pattern
left behind.

```{r lucas-resid, eval = has_map, fig.alt = "Map of mixture-model residuals across Lucas County"}
ggplot(lucas_sf) +
  geom_sf(aes(colour = resid), size = 0.6) +
  scale_colour_gradient2(low = "#2c7bb6", mid = "grey85", high = "#d7191c") +
  labs(title = "Mixture-model residuals", colour = "resid")
```

For reporting, the spatial-block bootstrap (`variance = "boot"`) carries the regime
labels through resampling; the `vignette("spmixqr-mixtures")` covers the gate and
slope-surface machinery in full.

## 11. Diagnostics: can you trust it?

```{r diag}
d <- fit$diagnostics
c(converged = d$converged,
  moran_before_p = round(mr$before$p_value, 3),
  moran_after_p  = round(mr$after$p_value, 3))
```

* **Did the spatial term help?** Compare residual Moran's I before and after (Section 8).
  If it was already non-significant, there is little spatial structure to absorb: the
  "negative control" case.
* **Inference.** Report the **spatial-block bootstrap**, not the sandwich:
  `spmixqr(..., variance = "boot")` resamples connected blocks of the graph and refits.
* **Weak identifiability.** The CAR strength `car_alpha` is held fixed (default 0.95), and
  $\lambda_\phi$ is selectable by BIC via `spmixqr_select()`. Neither is a precisely
  estimable variance, so we do not over-interpret their magnitudes.
* **Small samples.** Columbus has 49 units. The point estimates are stable, but spatial
  variance parameters are weakly identified at this size; lean on the bootstrap and avoid
  many regimes.

## 12. Practical guidance and pitfalls

* **Areal CAR fits use `spatial_coef = FALSE`** and pass the region id as `coords`. The
  CAR term carries the spatial level; the slope-surface machinery (`spatial_coef = TRUE`)
  is the separate point-data feature in `vignette("spmixqr-mixtures")`.
* **Choosing `W`.** Queen contiguity is the default for polygons; rook is stricter
  (shared edges only); for point data, use distance bands or k-NN. Check the substantive
  conclusions against a second `W`.
* **Gate vs CAR.** A free spatial *gate* and a CAR *intercept* are not separately
  identified. Setting `spatial_error = TRUE` turns the spatial gate off, and an explicit
  `spatial_gate = TRUE` raises a guardrail error. A covariate gate (`gating = ~ z`) is
  fine.
* **Regimes.** `G > 1` gives a mixture of spatial quantile regressions (latent
  neighbourhood types); use it only with enough units.

## 13. How `spmixqr` relates to other tools

`quantreg` fits quantile regression with no spatial structure; `spatialreg` and `spdep`
fit spatial autoregressive *mean* models; `CARBayes` fits CAR models but not a quantile
likelihood. The only dedicated spatial quantile package, `McSpatial`, has been archived
since 2021. `spmixqr` provides a maintained spatial-error quantile regression with
first-class queen, rook, distance, k-NN, and user-supplied weights, a Moran's-I
diagnostic, and the mixture extension [@reich2011].

## References
