---
title: "Mixtures and spatially varying effects (gate and slope surfaces)"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: ../spmixqr.bib
vignette: >
  %\VignetteIndexEntry{Mixtures and spatially varying effects (gate and slope surfaces)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4.2, dpi = 96)
has_gg <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r setup}
library(spmixqr)
set.seed(1)
```

This primer is for a researcher who has spatial data and a distributional question
but no prior exposure to spatial quantile mixtures. It walks from the question to a
reproducible report, using a built-in version of the Meuse river soil dataset.

## 1. The question

Soil zinc near the river Meuse is contaminated unevenly. Two questions a mean
regression cannot answer:

* How does a **quantile** of contamination (here the median, though any `tau` works,
  including the upper tail) respond to distance from the river, and does that
  response **change across the flood plain**?
* Are there latent **regimes**, patches that behave like a different contamination
  process, and where are they?

`spmixqr` answers both at once. It splits the area into latent regimes whose
membership drifts smoothly over space, fits a quantile regression in each, and lets
each regime's covariate effect be a spatial surface.

## 2. The model in one page

For a chosen quantile level $\tau$ (say the 0.9 upper tail), each observation belongs
to one of $G$ latent regimes. The probability of regime $k$ at location $s$ is a
smooth spatial surface, a softmax of a spatial basis (the *gate*). Within regime
$k$, the $\tau$-th conditional quantile of the response is a linear predictor whose
slopes vary over space:
$$ Q_\tau(y \mid \text{regime }k, x, s) = \beta_{k0} + \sum_{j} x_j\,\beta_{kj}(s). $$
The intercept $\beta_{k0}$ is a scalar. Spatial level variation is carried by the
gate, since a free spatial intercept surface is not separately identified from it.
Estimation is a penalised EM that reuses the `mixqr` quantile-regression core
[@wuyao2016] for the components and a spatially penalised multinomial-logit gate
[@fernandezgreen2002; @beraha2021]. The component step uses a convolution-smoothed
check loss [@fgh2021; @tanwangzhou2022] so a roughness penalty can be applied. The
single-regime case is a penalised spatially-varying-coefficient quantile regression
in the lineage of @reich2011.

## 3. The data

```{r data}
data(meuse_zinc)
meuse_zinc$lzinc <- log(meuse_zinc$zinc)
str(meuse_zinc)
```

`lzinc` is log zinc (ppm); `dist` is normalised distance to the river; `x`, `y` are
coordinates. Zinc is right-skewed and decays with distance, but the *rate* of decay
plausibly differs across the plain.

## 4. Fitting your first model

We model the conditional median ($\tau = 0.5$) of log zinc as a function of distance,
with two spatial regimes. Refit with `tau = 0.9` to study the upper tail.

```{r fit}
fit <- spmixqr(lzinc ~ dist, data = meuse_zinc, coords = c("x", "y"),
               G = 2, tau = 0.5, variance = "sandwich",
               control = spmixqr_control(nstart = 5L, k = 15L, seed = 1))
fit
```

The fit reports a `label_stability` of about 0.26: the two regimes' distance-effect
surfaces cross at roughly a quarter of locations, so the global "regime 1 / regime 2"
labels are coherent across most, but not all, of the plain. Section 9 returns to this;
read the maps below with that caveat in mind.

## 5. Interpreting the estimates

```{r summary}
summary(fit)
```

Each regime has its own intercept (overall contamination level) and an average
distance slope. A steep negative slope means zinc falls quickly away from the river;
a flat slope means a regime where distance barely matters (persistent contamination).
The gate block reports any gating-covariate effects; here the spatial field does the
work, so read the gate as a *map* (next section), not a coefficient.

## 6. Getting the uncertainty right

Two cautions are built in. The asymmetric-Laplace likelihood is a *working*
likelihood, and the sandwich standard errors above are *classification-conditional*:
they treat the regime labels as known, so they tend to be optimistic. For reporting,
prefer the spatial-block bootstrap, which refits the whole pipeline and aligns labels
across replicates:

```{r boot, eval = FALSE}
fit_boot <- spmixqr(lzinc ~ dist, data = meuse_zinc, coords = c("x", "y"),
                    G = 2, tau = 0.5, variance = "boot",
                    control = spmixqr_control(nstart = 5L, k = 15L, seed = 1,
                                              boot_B = 200L, boot_block = 3L))
```

## 7. The headline question, answered

Does the distance effect on the median vary across space? The slope surface
answers it directly.

```{r slope-surface}
cs <- coef_surface(fit, covariate = 1)
tapply(cs$slope, cs$regime, function(s) round(range(s), 2))
```

A wide slope range means the distance effect is genuinely spatial within that regime,
not a single number.

## 8. Reading the fit in pictures

```{r maps, eval = has_gg, fig.alt = "Gate and slope-surface maps"}
library(ggplot2)
theme_set(theme_minimal(base_size = 11))

gs <- gate_surface(fit)
ggplot(subset(gs, regime == "2"), aes(coord1, coord2, colour = prob)) +
  geom_point(size = 2) +
  scale_colour_viridis_c(limits = c(0, 1)) +
  labs(title = "Where regime 2 dominates", x = "easting", y = "northing",
       colour = "P(regime 2)")
```

```{r slope-map, eval = has_gg, fig.alt = "Slope surface map"}
ggplot(cs, aes(coord1, coord2, colour = slope)) +
  geom_point(size = 2) + facet_wrap(~ regime) +
  scale_colour_viridis_c(option = "magma") +
  labs(title = "Distance-effect surface by regime", x = "easting", y = "northing")
```

## 9. Diagnostics: can you trust it?

```{r diag}
d <- fit$diagnostics
c(converged = d$converged, starts = d$n_starts,
  gate_cond = round(d$gate_cond, 1),
  class_entropy = round(d$class_entropy, 2),
  label_stability = round(d$label_stability, 2))
```

`class_entropy` near 0 means confident classification; near 1 means overlapping
regimes. `label_stability` is the fraction of locations where the regime slope
surfaces cross: a high value warns that a single global label is only approximately
coherent across space (a known v1 limitation). Here it sits near 0.26.

**A negative control.** When there is no spatial structure, the method should report
none. We simulate a constant gate and flat slopes, then check that the gate surface
stays flat:

```{r negctrl}
nc <- sim_spmixqr(n = 200, gate_slope = 0, coef_slope = 0, seed = 7)
fit_nc <- spmixqr(y ~ x, nc$data, coords = nc$coords, G = 2, tau = 0.5,
                  lambda_gate = 10, lambda_coef = 10, variance = "none",
                  control = spmixqr_control(nstart = 4L, seed = 1))
g_nc <- gate_surface(fit_nc)
round(sd(g_nc$prob[g_nc$regime == "2"]), 3)   # small => no manufactured structure
```

## 10. Practical guidance and pitfalls

* **Keep `G` small** (2--3). Spatial mixtures are data-hungry, and label coherence
  degrades as regimes proliferate.
* **Standardise coordinates.** This is handled internally, but supply coordinates on
  a sensible common scale.
* **Choose the smoothing** with `spmixqr_select()` (BIC or cross-validated check
  loss) rather than guessing `lambda_gate`/`lambda_coef`.
* **Report bootstrap intervals**, not the sandwich, for inference.
* Watch the `label_stability` warning: crossing slope surfaces mean the regime
  identities are not globally fixed.

## 11. How `spmixqr` relates to other tools

`quantreg` fits one quantile regression with no spatial structure; `mixqr` adds
latent regimes but fixed mixing; `mixqrgate` lets mixing depend on covariates.
`spmixqr` makes the mixing *spatial* and the component effects *spatial surfaces*. No
maintained R package fits spatial quantile regression (`BSquare` and `McSpatial` are
archived), so the single-regime case fills that gap too [@reich2011].

## 12. Reporting and reproducibility

```{r report}
list(tau = fit$tau, G = fit$G,
     intercepts = round(fit$beta_const[1, ], 2),
     avg_slopes = round(fit$beta_const[2, ], 2),
     bic = round(fit$bic, 1))
```

A spatial quantile mixture turns "how does contamination respond to distance" into a
*map* of regime-specific tail responses. Through the diagnostics and the negative
control, it declines to invent structure the data do not support.

## References
