---
title: "Spatial CAR-error quantile regression (areal and point data)"
output: rmarkdown::html_vignette
bibliography: ../spmixqr.bib
vignette: >
  %\VignetteIndexEntry{Spatial CAR-error quantile regression (areal and point data)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4)
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r setup}
library(spmixqr)
set.seed(1)
```

## What the CAR spatial-error term adds

`spmixqr` v0.1.0 modelled spatial structure through an `mgcv` basis on the **gate**
and **slope surfaces**. Version 0.2.0 adds a weights-matrix-driven alternative: each
regime carries a per-unit **conditional-autoregressive (CAR)** spatial random effect
`phi_k`, a mean-zero spatial-level surface with a Leroux proper precision

$$Q(\alpha) = \alpha (D - W) + (1-\alpha) I, \qquad \alpha \in [0,1],$$

which recovers the intrinsic CAR (ICAR/Besag) at $\alpha = 1$ and independence at
$\alpha = 0$ [@leroux2000]. The asymmetric-Laplace working likelihood of the quantile
components makes a Gaussian CAR effect the right penalty for the M-step
[@kozumi2011]. We solve the penalised check loss directly, so $\lambda_\phi$ is a
tuning parameter chosen by BIC. With `G = 1` this is a single-population spatial-error
quantile regression; with `G > 1` it is a mixture of spatial quantile regressions.

**Identification.** A free spatial gate aliases with the CAR intercept surface, so
`spatial_error = TRUE` turns the spatial gate off by default, and an explicit
`spatial_gate = TRUE` then raises a guardrail error. A per-connected-component
**sum-to-zero** constraint on `phi_k` is absorbed into the incidence and precision (as
`mgcv`'s mrf basis does via `absorb.cons`), so the intercept carries the level and the
CAR surface is a mean-zero deviation. `alpha` is **fixed** (default 0.95) and disclosed
as weakly identified; the check loss does not select it (verified degenerate).

## Building the weights matrix

`spq_weights()` constructs a symmetric, sparse `W` from polygons (queen or rook
contiguity), point coordinates (distance band or k-nearest-neighbour, symmetrised), an
`spdep` `nb`/`listw`, or a user-supplied matrix. The default `style = "B"` (symmetric
binary) is required for a valid Gaussian Markov random field precision.

```{r weights}
data(nc_sids_W)
W <- spq_weights(nc_sids_W, type = "supplied")
W
```

## A census-tract areal example: North Carolina SIDS

The shipped `nc_sids` data are sudden-infant-death counts for the 100 North Carolina
counties (1974--78), the canonical areal/lattice dataset [@cressie1993]. We model the
log SIDS rate against the proportion of non-white births, with a county-level CAR
spatial-error surface.

```{r nc-fit}
data(nc_sids)
fit <- spmixqr(log_sids ~ pnw, nc_sids, coords = nc_sids$county,
               G = 1, tau = 0.5, spatial_error = TRUE, spatial_coef = FALSE,
               spatial_W = W, variance = "none",
               control = spmixqr_control(nstart = 1L, seed = 1))
coef(fit)
```

The fitted CAR surface is a tidy data frame for mapping:

```{r nc-phi}
ph <- phi_surface(fit)
head(ph)
```

```{r nc-map, eval = have_ggplot, fig.alt = "CAR spatial-error surface by county"}
library(ggplot2)
ph$east  <- nc_sids$east[match(ph$unit, nc_sids$county)]
ph$north <- nc_sids$north[match(ph$unit, nc_sids$county)]
ggplot(ph, aes(east, north, colour = phi)) +
  geom_point(size = 3) +
  scale_colour_gradient2() +
  labs(title = "County CAR spatial-error surface (median regression)",
       x = "easting", y = "northing", colour = expression(phi)) +
  theme_minimal()
```

### Residual spatial autocorrelation (a negative control)

`moran_resid()` runs a permutation Moran's I on the responsibility-weighted residual
aggregated to the unit, before and after the CAR term. The `nc_sids` log-rate residuals
carry **little spatial signal to begin with**: the before-CAR Moran's I is small and
non-significant. This is a negative control rather than a power demonstration. With
nothing for the CAR term to absorb, BIC drives `lambda_phi` toward a near-flat surface.

```{r nc-moran}
moran_resid(fit, nsim = 199)
```

The before-significant to after-non-significant story appears below, on a simulated DGP
that *does* carry genuine residual autocorrelation.

## A point example: meuse zinc

For point data, `spq_weights()` builds the graph from coordinates (here a symmetrised
k-nearest-neighbour graph). Each distinct location is its own unit.

```{r meuse}
data(meuse_zinc)
cc <- as.matrix(meuse_zinc[, c("x", "y")])
Wp <- spq_weights(cc, type = "knn", k = 6)
fit_p <- spmixqr(log(zinc) ~ dist, meuse_zinc, coords = cc, G = 1, tau = 0.5,
                 spatial_error = TRUE, spatial_coef = FALSE, spatial_W = Wp,
                 variance = "none", control = spmixqr_control(nstart = 1L, seed = 1))
coef(fit_p)
```

## Recovery on a known DGP

`sim_spmixqr(spatial_error = TRUE)` generates lattice data with a known CAR effect, so
we can check recovery directly.

```{r recovery}
d <- sim_spmixqr(n = 500, G = 1, tau = 0.5, spatial_error = TRUE,
                 lattice = 7, car_rho = 1.5, seed = 7)
f <- spmixqr(y ~ x, d$data, coords = d$region, G = 1, tau = 0.5,
             spatial_error = TRUE, spatial_coef = FALSE, spatial_W = d$spatial_W,
             variance = "none", control = spmixqr_control(nstart = 1L, seed = 1))
cor(f$car$phi[, 1], d$truth$phi[, 1])    # phi recovery
sum(f$car$phi[, 1])                       # mean-zero constraint (exact)
```

### The CAR term absorbs genuine residual autocorrelation

Here the unobserved CAR surface induces real spatial autocorrelation in the residuals,
so `moran_resid()` shows the intended power story: Moran's I is large and significant
*before* the CAR term and drops to non-significance *after* it.

```{r recovery-moran}
moran_resid(f, nsim = 199)
```

## Inference and caveats

The asymmetric-Laplace working likelihood makes naive standard errors optimistic. We
recommend reporting the **spatial-block bootstrap**, which for areal data resamples
contiguous connected-component blocks of `W`; an i.i.d. fallback would destroy the
spatial dependence. The bootstrap is the cost centre, so the example below is a small
smoke run.

```{r boot, eval = FALSE}
## \donttest: refit x B is slow; use moderate L / few reps for the block bootstrap.
fit_b <- spmixqr(log_sids ~ pnw, nc_sids, coords = nc_sids$county, G = 1, tau = 0.5,
                 spatial_error = TRUE, spatial_coef = FALSE, spatial_W = W,
                 variance = "boot",
                 control = spmixqr_control(nstart = 1L, seed = 1, boot_B = 50L))
vcov(fit_b)$note
```

Caveats, all disclosed in `summary()`: `alpha` is weakly identified (fixed, not
estimated); `lambda_phi` is a penalty, not a precise variance; the block bootstrap
under-covers with few blocks [@lahiri2003]; and the spatial-**lag** ($\rho W y$) model
is deferred (v1 models autocorrelated unobservables, not spillovers).

## References
