---
title: "Spatial CAR-error quantile regression (areal and point data)"
output: rmarkdown::html_vignette
bibliography: ../spmixqr.bib
vignette: >
  %\VignetteIndexEntry{Spatial CAR-error quantile regression (areal and point data)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4)
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r setup}
library(spmixqr)
set.seed(1)
```

## What the CAR spatial-error term adds

`spmixqr` v0.1.0 modelled spatial structure through an `mgcv` basis on the **gate**
and **slope surfaces**. Version 0.2.0 adds a weights-matrix-driven alternative: each
regime carries a per-unit **conditional-autoregressive (CAR)** spatial random effect
`phi_k`, a mean-zero spatial-level surface with a Leroux proper precision

$$Q(\alpha) = \alpha (D - W) + (1-\alpha) I, \qquad \alpha \in [0,1],$$

which recovers the intrinsic CAR (ICAR/Besag) at $\alpha = 1$ and independence at
$\alpha = 0$ [@leroux2000]. The asymmetric-Laplace working likelihood of the quantile
components makes a Gaussian CAR effect the right penalty for the M-step
[@kozumi2011]. We solve the penalised check loss directly, so $\lambda_\phi$ is a
tuning parameter chosen by BIC. With `G = 1` this is a single-population spatial-error
quantile regression; with `G > 1` it is a mixture of spatial quantile regressions.

**Identification.** A free spatial gate aliases with the CAR intercept surface, so
`spatial_error = TRUE` turns the spatial gate off by default, and an explicit
`spatial_gate = TRUE` then raises a guardrail error. A per-connected-component
**sum-to-zero** constraint on `phi_k` is absorbed into the incidence and precision (as
`mgcv`'s mrf basis does via `absorb.cons`), so the intercept carries the level and the
CAR surface is a mean-zero deviation. `alpha` is **fixed** (default 0.95) and disclosed
as weakly identified; the check loss does not select it (verified degenerate).

## Building the weights matrix

`spq_weights()` constructs a symmetric, sparse `W` from polygons (queen or rook
contiguity), point coordinates (distance band or k-nearest-neighbour, symmetrised), an
`spdep` `nb`/`listw`, or a user-supplied matrix. The default `style = "B"` (symmetric
binary) is required for a valid Gaussian Markov random field precision.

```{r weights}
data(nc_sids_W)
W <- spq_weights(nc_sids_W, type = "supplied")
W
```

## A census-tract areal example: North Carolina SIDS

The shipped `nc_sids` data are sudden-infant-death counts for the 100 North Carolina
counties (1974--78), the canonical areal/lattice dataset [@cressie1993]. We model the
log SIDS rate against the proportion of non-white births, with a county-level CAR
spatial-error surface.

```{r nc-fit}
data(nc_sids)
fit <- spmixqr(log_sids ~ pnw, nc_sids, coords = nc_sids$county,
               G = 1, tau = 0.5, spatial_error = TRUE, spatial_coef = FALSE,
               spatial_W = W, variance = "none",
               control = spmixqr_control(nstart = 1L, seed = 1))
coef(fit)
```

The fitted CAR surface is a tidy data frame for mapping:

```{r nc-phi}
ph <- phi_surface(fit)
head(ph)
```

```{r nc-map, eval = have_ggplot, fig.alt = "CAR spatial-error surface by county"}
library(ggplot2)
ph$east  <- nc_sids$east[match(ph$unit, nc_sids$county)]
ph$north <- nc_sids$north[match(ph$unit, nc_sids$county)]
ggplot(ph, aes(east, north, colour = phi)) +
  geom_point(size = 3) +
  scale_colour_gradient2() +
  labs(title = "County CAR spatial-error surface (median regression)",
       x = "easting", y = "northing", colour = expression(phi)) +
  theme_minimal()
```

### Residual spatial autocorrelation (a negative control)

`moran_resid()` runs a permutation Moran's I on the responsibility-weighted residual
aggregated to the unit, before and after the CAR term. The `nc_sids` log-rate residuals
carry **little spatial signal to begin with**: the before-CAR Moran's I is small and
non-significant. This is a negative control rather than a power demonstration. With
nothing for the CAR term to absorb, BIC drives `lambda_phi` toward a near-flat surface.

```{r nc-moran}
moran_resid(fit, nsim = 199)
```

The before-significant to after-non-significant story appears below, on a simulated DGP
that *does* carry genuine residual autocorrelation.

## A point example: meuse zinc

For point data, `spq_weights()` builds the graph from coordinates (here a symmetrised
k-nearest-neighbour graph). Each distinct location is its own unit.

```{r meuse}
data(meuse_zinc)
cc <- as.matrix(meuse_zinc[, c("x", "y")])
Wp <- spq_weights(cc, type = "knn", k = 6)
fit_p <- spmixqr(log(zinc) ~ dist, meuse_zinc, coords = cc, G = 1, tau = 0.5,
                 spatial_error = TRUE, spatial_coef = FALSE, spatial_W = Wp,
                 variance = "boot",
                 control = spmixqr_control(nstart = 1L, seed = 1, boot_B = 50L))
coef(fit_p)
```

Because the outcome is `log(zinc)`, the spatial effect is most interpretable on the
multiplicative scale: `phi_surface(scale = "exp")` reports `exp(phi)`, the factor by
which zinc runs above or below what distance-to-river predicts. With `ci = TRUE` we keep
only the locations where that factor is reliably away from 1.

```{r meuse-exp}
pm <- phi_surface(fit_p, ci = TRUE, scale = "exp")
range(round(pm$mult, 2))                       # e.g. 1.18 = ~18% above expected
sprintf("%d of %d sites are credible (zinc reliably above/below expectation)",
        sum(pm$credible), nrow(pm))
```

## NNGP: a weights-matrix-free spatial error for point data

For point data, a contiguity/k-NN `W` is an ad hoc discretisation of a continuous
surface. A **nearest-neighbour Gaussian process** (NNGP; Datta, Banerjee, Finley &
Gelfand 2016) instead places a continuous Matern field on the points and represents it by
a *sparse* precision `Q = (I - B)' D^{-1} (I - B)` built from each point's `m` nearest
neighbours. `spq_weights(type = "nngp")` returns that precision; it plugs into the same
penalised spatial-error M-step as the CAR term — but it is `W`-free, scalable (the
precision is sparse), and continuous-domain. Because the precision is proper (full rank),
`phi` is fit unconstrained (no sum-to-zero).

```{r nngp}
cc <- as.matrix(meuse_zinc[, c("x", "y")])
Wn <- spq_weights(cc, type = "nngp", m = 10, nu = 0.5)   # 10-neighbour exponential field
Wn
fit_n <- spmixqr(log(zinc) ~ dist, meuse_zinc, coords = cc, G = 1, tau = 0.5,
                 spatial_error = TRUE, spatial_coef = FALSE, spatial_W = Wn,
                 variance = "none", control = spmixqr_control(nstart = 1L, seed = 1))
coef(fit_n)
```

The Matern `range` and smoothness `nu` are weakly identified from one realisation (Zhang
2004), so they are fixed (or chosen by BIC over a few `spq_weights(range = ...)` objects),
exactly as the roughness penalty is. For point GPs use `variance = "boot"`: the
classification-conditional sandwich forms a dense per-unit covariance.

## Spatial+: guarding against spatial confounding

Adding *any* spatial random effect — CAR or NNGP — can bias the covariate effects when a
covariate itself varies smoothly over space: the spatial term and the covariate compete
for the same variation, attenuating (or even flipping) the slope (Hodges & Reich 2010).
The sound remedy is **Spatial+** (Dupont, Wood & Augustin 2022): residualise each
covariate against a spatial smooth and fit on the residuals, so the covariate keeps only
its *non-spatial* variation. It is loss-agnostic, so it composes with the quantile EM. (We
do **not** offer restricted spatial regression, which the recent literature has refuted as
a general fix.)

`sim_spatial_confound()` makes the problem concrete: a covariate `x` confounded with a
spatial field that also drives `y`, with a known true slope of 1.

```{r splus}
d <- sim_spatial_confound(n = 400, beta = 1.0, confound = 0.9, seed = 11)
Wn2 <- spq_weights(d$coords, type = "nngp", m = 10)
ctl <- spmixqr_control(nstart = 1L, seed = 1)
b <- function(f) round(coef(f)["x", 1], 3)
f_naive <- spmixqr(y ~ x, d$data, coords = d$coords, G = 1, tau = 0.5,
                   spatial_error = FALSE, spatial_coef = FALSE, variance = "none", control = ctl)
f_nngp  <- spmixqr(y ~ x, d$data, coords = d$coords, G = 1, tau = 0.5, spatial_error = TRUE,
                   spatial_coef = FALSE, spatial_W = Wn2, variance = "none", control = ctl)
f_splus <- spmixqr(y ~ x, d$data, coords = d$coords, G = 1, tau = 0.5, spatial_error = TRUE,
                   spatial_coef = FALSE, spatial_W = Wn2, spatial_plus = TRUE,
                   variance = "none", control = ctl)
c(truth = 1, naive = b(f_naive), nngp_only = b(f_nngp), spatial_plus = b(f_splus))
```

The naive slope is badly inflated; the NNGP term absorbs part of the confounding but, being
penalised, leaves a residual bias; **Spatial+** removes most of what remains. The catch
(Frisch–Waugh–Lovell): Spatial+ deconfounds only to the extent its residualisation smooth
*out-resolves* the penalised spatial term — residualising on the very same basis is a
no-op. The residualisation is mean-based, so it cleanly deconfounds the quantile slope for
symmetric error; the reported slopes are the effect of the non-spatial part of each
covariate (see `summary(f_splus)`).

### Identification under the check loss (why mean-based residualisation works)

Spatial+ was derived for least-squares (mean) regression, so does a *mean*-based
residualisation deconfound the *quantile* slope? Yes — and at every quantile at once.
Under an additive model `y = m(s) + beta_x x + eps` with an exogenous error
`eps` and `x = E[x | s] + e`, removing `E[x | s]` makes the conditional quantile affine in
the residual with slope `beta_x` *for every* `tau`:
`Q_tau(y | r, s) = beta_x r + h_tau(s)`, where `h_tau(s)` is absorbed by the retained
spatial term. The key point is that deconfounding is a conditional-orthogonality property
— it does not depend on `tau` — so a single least-squares residualisation works across the
whole distribution; there is no need to residualise separately per quantile. (A full proof
and its boundary cases are in the package's methodological notes.) The recovery holds in
the tails, not just at the median:

```{r splus-tau}
taus <- c(0.1, 0.5, 0.9)
rbind_tau <- function(tau) {
  fn <- spmixqr(y ~ x, d$data, coords = d$coords, G = 1, tau = tau, spatial_error = FALSE,
                spatial_coef = FALSE, variance = "none", control = ctl)
  fp <- spmixqr(y ~ x, d$data, coords = d$coords, G = 1, tau = tau, spatial_error = TRUE,
                spatial_coef = FALSE, spatial_W = Wn2, spatial_plus = TRUE,
                variance = "none", control = ctl)
  c(tau = tau, naive = b(fn), spatial_plus = b(fp))
}
round(do.call(rbind, lapply(taus, rbind_tau)), 3)   # true slope = 1 at every tau
```

The two practical conditions are that the covariate enters additively (so a single slope
exists) and that the residualisation smooth, together with the spatial term, out-resolves
the covariate's spatial mean (the Frisch–Waugh–Lovell gap above).

## Recovery on a known DGP

`sim_spmixqr(spatial_error = TRUE)` generates lattice data with a known CAR effect, so
we can check recovery directly.

```{r recovery}
d <- sim_spmixqr(n = 500, G = 1, tau = 0.5, spatial_error = TRUE,
                 lattice = 7, car_rho = 1.5, seed = 7)
f <- spmixqr(y ~ x, d$data, coords = d$region, G = 1, tau = 0.5,
             spatial_error = TRUE, spatial_coef = FALSE, spatial_W = d$spatial_W,
             variance = "none", control = spmixqr_control(nstart = 1L, seed = 1))
cor(f$car$phi[, 1], d$truth$phi[, 1])    # phi recovery
sum(f$car$phi[, 1])                       # mean-zero constraint (exact)
```

### The CAR term absorbs genuine residual autocorrelation

Here the unobserved CAR surface induces real spatial autocorrelation in the residuals,
so `moran_resid()` shows the intended power story: Moran's I is large and significant
*before* the CAR term and drops to non-significance *after* it.

```{r recovery-moran}
moran_resid(f, nsim = 199)
```

## Inference and caveats

The asymmetric-Laplace working likelihood makes naive standard errors optimistic. We
recommend reporting the **spatial-block bootstrap**, which for areal data resamples
contiguous connected-component blocks of `W`; an i.i.d. fallback would destroy the
spatial dependence. The bootstrap is the cost centre, so the example below is a small
smoke run.

```{r boot, eval = FALSE}
## \donttest: refit x B is slow; use moderate L / few reps for the block bootstrap.
fit_b <- spmixqr(log_sids ~ pnw, nc_sids, coords = nc_sids$county, G = 1, tau = 0.5,
                 spatial_error = TRUE, spatial_coef = FALSE, spatial_W = W,
                 variance = "boot",
                 control = spmixqr_control(nstart = 1L, seed = 1, boot_B = 50L))
vcov(fit_b)$note
```

Caveats, all disclosed in `summary()`: `alpha` is weakly identified (fixed, not
estimated); `lambda_phi` is a penalty, not a precise variance; the block bootstrap
under-covers with few blocks [@lahiri2003]; and the spatial-**lag** ($\rho W y$) model
is deferred (v1 models autocorrelated unobservables, not spillovers).

## References
